---
title: "A quick overview of CCLE data"
title-block-banner: "#000061"
subtitle: ""
description: ""
date: last-modified
date-format: "YYYY-MM-DD"
mainfont: Figtree
sansfont: Georgia
footnotes-hover: true
reference-location: margin
lang: en
number-sections: false
crossref:
  chapters: true
author:
  - name: Mattias Vesterlund
    orcid: 0000-0001-9471-6592
    email: matves29@kth.se
    url: https://kth.se
    affiliations:
      - name: KTH Library
        address: Osquars backe 30
        city: Stockholm
        postal-code: 100 44
        url: https://www.kth.se/biblioteket
highlight-style: pygments
fig-cap-location: top
format: 
  kth-quarto-html: default
editor: visual
bibliography: references.bib
link-citations: true
csl: IEEE.csl
---

### Abstract

This is a brief example of a few exploratory plots for comparing expression patterns for genes between a few cell lines from the cancer cell line Encyclopedia (CCLE).

#### A QC plot to check the data

Here is a boxplot of expression levels as an example of a QC-plot.

```{r, warning=F, echo=F,message=F, dev = "png", dev.args=list(bg="transparent")}
library(ggplot2)
library(ktheme)
library(matrixStats)
library("ggplot2")
library("reshape2")
library("gplots")
library("devtools")
library("plotly")
library("wesanderson")
library("msigdbr")
library("ggbeeswarm")
library("ggpubr")
library("ggridges")
library("ggExtra")
library("robustbase")
library("ggpubr")
library("survival")
library("rms")
library("survminer")
library("factoextra")
library("NbClust")
library("survival")
library("survminer")
library("gtsummary")
library("rms")
library("tidyr")
library("reshape2")
library("gt")
library(kableExtra)
select <- dplyr::select

#read CCLE gene expression data 
CCLEmatrix <- as.matrix(read.table("data/CCLEmatrix.txt",header=T, row.names=1, check.names = FALSE, na.strings=c("", "NA", "Inf"), sep="\t"))

#Also some metadata about the observations
CCLEinfo <- read.table("data/CCLEinfo.txt",header=T, row.names=1, check.names = FALSE, na.strings=c("", "NA", "Inf"), sep="\t")

CCLEinfo$Type<-replace(CCLEinfo$Type, CCLEinfo$Type == "HAEMATOPOIETIC_AND_LYMPHOID_TISSUE", "BLOOD")
CCLEinfo$Type<-replace(CCLEinfo$Type, CCLEinfo$Type == "CENTRAL_NERVOUS_SYSTEM", "CNS")

Cluster1<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("BREAST")),])

Cluster2<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("SKIN")),])

Cluster3<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("LUNG_SCLC")),])

Cluster4<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("LUNG_NSCLC")),])

Cluster5<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("BLOOD")),])

Cluster6<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("CNS")),])

testlines<-c(Cluster1[1:7],Cluster6[1:7],Cluster5[1:7],Cluster4[1:7],Cluster3[1:7],Cluster2[1:7])

ccle_melt<-melt(CCLEmatrix[,testlines])
ccle_melt$type<-CCLEinfo[as.character(ccle_melt$Var2),2]



ggplot(ccle_melt) + geom_boxplot(aes(y=value,group = Var2 , fill = type )) +
  scale_color_kth() +
  scale_fill_kth()+
  theme_kth_neo() +
  theme(legend.position = "top",
        axis.line = element_line(color='black'),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())


```

There is quite a bit of variability between different lineages. But overall it looks ok. We can't assume all cell types have the same overall level of expression. To get an overview of how individual cell lines cluster and differ we can run a tSNE analysis to visualize the dataset in two dimensions. 

```{r, warning=F, echo=F,message=F, dev = "png", dev.args=list(bg="transparent")}
library(Rtsne)
tsne_out <- Rtsne(t(CCLEmatrix))

tsne_plot <- data.frame(x = tsne_out$Y[,1], 
                        y = tsne_out$Y[,2])
tsne_plot$Type<-CCLEinfo$Type

ggplot(tsne_plot, fill=Type) + geom_point(aes(x=x,y=y,color=Type)) +
  scale_color_kth() +
  scale_fill_kth()+
  theme_kth_neo() +
  theme(legend.position = "top",
        axis.line = element_line(color='black'),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank())


```

#### Interactivity as an example of exploration

Here we add an example of how interactivity can be used in data exploration using the CCLE dataset. Here is a comparison of Breast Cancer cell lines and Skin Cancer cell lines. To figure out which genes are the most different we can hover over the data points.

```{r, warning=F, echo=F,message=F, dev = "png", dev.args=list(bg="transparent")}



Cluster1<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("BREAST")),])

Cluster2<-rownames(CCLEinfo[which(CCLEinfo$Type %in% c("SKIN")),])


skin_vs_breast_matrix <- CCLEmatrix[,c(Cluster1,Cluster2)]

#First fold change
C1vsC2Fold <- rowMeans(skin_vs_breast_matrix[,Cluster1])-rowMeans(skin_vs_breast_matrix[,Cluster2])

#run a t.test by row
t.result <- apply(skin_vs_breast_matrix, 1, 
                  function (x) t.test(x[Cluster1],x[Cluster2]))

#get the p-values
p_values <- unlist(lapply(t.result, 
                          function(x) x$p.value))

#correct for multiple testing
adj.p_values <- p.adjust(p_values, method = "fdr")

#Results Table
SvB_DE_results <- data.frame(fc=C1vsC2Fold, 
                            p.value=p_values, 
                            adj.p.value=adj.p_values,
                            row.names=rownames(skin_vs_breast_matrix), 
                            genesymbol = rownames(skin_vs_breast_matrix))

#results table sorted by adjusted p-value
SvB_DE_results_sorted <- SvB_DE_results[order(SvB_DE_results$adj.p.value,
                                            decreasing=FALSE),]

#here we make a standard ggplot object, but we also add text as an additional aesthetic.
p<-ggplot(SvB_DE_results, aes(fc,(-1)*log10(adj.p.value),size=1.5,col=((-1)*log10(adj.p.value)),text=paste0("Gene:  ",genesymbol)))+
  geom_point()+theme_bw()+
  labs(y= "-log10(adjusted p-value)", x = "LogFC",legend="-log10(adjusted p-value)")+
  scale_color_gradientn(name="neglog10 P-value",colours =palette_kth_neo(n=4,type="seq"))+
  ggtitle("Volcano plot \nFDR corrected ttest on CCLE data") +
  theme(plot.title = element_text(size = 9))

#then we call plotly and tell it to use the text from the aesthetic in the ggobj as a tooltip.
ggplotly(p,tooltip = "text")




```
